---
title: "1. Data Wrangling"
author: "Solveig Sølverød Mo and Eve Zeyl Fiskebeck"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
    pjs_df: "data/path.xlsx"
    pjs_sheet: "Cleaned_data" 
    vigas_df: "data/path.xlsx"
    source_code: 'code/R'
execute: 
  echo: false
  cache: false
format: html
knitr:
  opts_chunk: 
    collapse: true
    comment: "#>" 
    R.options:
      knitr.graphics.auto_pdf: true
editor_options: 
  chunk_output_type: console
---
<!-- Good to have at hand as comment 
- NVI Rstudio doc : <https://github.com/NorwegianVeterinaryInstitute/posit-team/wiki/Documentation>
- using Renv (allows to work on same versions) <https://rstudio.github.io/renv/articles/renv.html>
- Github repo : <https://github.com/NorwegianVeterinaryInstitute/2025_PAWH_APEC1>


renv::snapshot() # to run when new package installed
renv::load() # is the command to run when the other has updated the script and new packages are installed. 
This is now integrated in the .Rprofile
renv::restore() # to reinstall everything - only if problems
--> 

# Preliminary Setup 
```{r setup}
library(here)
library(tidyverse)
library(readxl)
library(dplyr)
```



<!-- comment or do not run this if using Rstudio (ask Eve for explanation) -->
```{r params for vscode}
# Only when using VScode (explain)
params <- list()
params$pjs_df <- "data/path.xlsx"
params$pjs_sheet <- "Cleaned_data" 
params$vigas_df <- "data/path.xlsx"
params$source_code <- 'code/R'

params$source_code
here::here(params$source_code)
```


```{r source reused functions}
fun_files_path <- list.files(
  here::here(params$source_code), 
  pattern = "*.R", 
  full.names = TRUE
  )
# sourcing all the functions 
sapply(fun_files_path,source,.GlobalEnv)
```


# Data import and merging
```{r merge files from VIGAS and PJS}
# Solveig trying to merge files from VIGAS and metadata from PJS:
# removing samples from 2015 and 2025, which should not be included in this study.

VIGAS_data <- read_xlsx("data/2025-7-2-APEC_all-All_Fields.xlsx") %>%
  dplyr::mutate(Year = substr(Sample,1,4)) %>%
  dplyr::filter(Year !="2015")%>%
  dplyr:: filter(Year !="2025")

PJS_data <- read_xlsx("data/Uttrekk_PJS_010725_SSM.xlsx") 
#799(?) samples to be included in this APEC study. Merge file from VIGAS with file from PJS:

APEC_all_merged <- 
  VIGAS_data %>%
  dplyr::left_join(PJS_data, by = c("Sample"="ID"))
#APEC_all_merged includes data for the 799 included samples.
```

```{r}
#Solveig trying to see how many isolates are missing data from PJS:
dim(PJS_data) # give you rows and columns from PJS - so 4548 isolates and 28 columns
str(PJS_data) # gives you the structure of the data, so you can see what is in each column
glimpse(PJS_data) # anotherway to gives you the structure of the data
```


```{r}
missing_PJS_data_artnavn <-
  APEC_all_merged %>%
  dplyr::select(`artnavn`) %>%
  dplyr::group_by(`artnavn`) %>%
  dplyr::summarise(antall=n())%>%
  dplyr::arrange(desc(antall))

#3 missing information regarding artnavn.

# alternative way to look at that -  so you know which ones
APEC_all_merged %>%
  dplyr::filter(is.na(`artnavn`))%>%
  dplyr::select(Sample)
```


- updating artsnavn for the 3 samples
```{r}
APEC_all_merged <- 
  APEC_all_merged %>%
  dplyr::mutate(`artnavn` = 
                  ifelse(is.na(`artnavn`), "Rowan Ranger", `artnavn`)) 
```

```{r}
#check if any are missing hensikt, indicate that data have not been extracted from PJS:
# missing_PJS_hensikt <- 
#   APEC_all_merged %>%
#   select(`hensiktkode`) %>%
#   group_by(`hensiktkode`) %>%
#   summarise(antall=n())%>%
#   arrange(desc(antall))
APEC_all_merged %>%
  filter(is.na(`hensiktkode`)) 

#None missing hensiktkode after updated PJS uttrekk. Save an output of APEC_all_merged to identify PJS numbers missing hensiktskode:

#write_delim(APEC_all_merged, "data/APEC_all_merged_070525.txt", delim = "\t")
#errors in PJS corrected. Also need to include another "hensikt" in the extraction of data from PJS. Asked Madelaine to fix it. Will run the code again when metadata are updated.
```

- Merging data MLST with Enterobase

```{r}
# recoding - to NA where  MLST is missing
glimpse(APEC_all_merged)
APEC_all_merged <- 
  APEC_all_merged %>%
  dplyr::mutate(`mlst/sequence_type` = 
                  ifelse(`mlst/sequence_type` == "-", NA, `mlst/sequence_type`)) 


# import enterobase
enterobase <- 
  read_xlsx("data/mlst_enterobase_juni25.xlsx", na = "ND")

# renaming columns, makes it an easy solution when you have to merge info ...
enterobase <-  
  enterobase %>%
  dplyr::rename(enterobase_ST = ST)

# completing APEC_all_merged 
APEC_all_merged_cleaned <- 
  APEC_all_merged %>%
  dplyr::left_join(enterobase, by = c("Sample" = "Name")) %>%
  # first check filtering if NA and check - after I comment
  # dplyr::select(Sample, `mlst/sequence_type`, enterobase_ST) %>%
  # dplyr::filter(is.na(`mlst/sequence_type`)) %>%
  # replacing NA in `mlst/sequence_type` with enterobase_ST
  dplyr::mutate(`mlst/sequence_type` = ifelse(is.na(`mlst/sequence_type`), enterobase_ST, `mlst/sequence_type`))%>% 
  # removing the extra column from enterobase
  dplyr::select(-enterobase_ST) %>%
  # Filtering out the 3 isolates where MLST did not work
  dplyr::filter(!is.na(`mlst/sequence_type`))

# new control :  799 - 3 = 796 isolates with MLST - ok
glimpse(APEC_all_merged_cleaned)    
dim(APEC_all_merged_cleaned)

length(unique(APEC_all_merged_cleaned$Sample)) # 796 isolates with MLST

# glimpse(enterobase)
# View(enterobase)
```




# Virulence genes overview

- import table virulence genes from Camila
```{r}
camila_virulence_genes <- 
  read_xlsx("data/Virulence_genes_APEC.xlsx")

glimpse(camila_virulence_genes)
```


```{r}
# Verifying length unique samples
APEC_all_merged_cleaned %>%
  select(Sample, `Virulence_Finder/VirulenceGenes:ProteinName`) %>%
  pull(Sample) %>% unique() %>% length()
  
# Creating the long table 
APEC_all_merged_cleaned_long <- 
  APEC_all_merged_cleaned %>%
  dplyr::rename(tosplit = `Virulence_Finder/VirulenceGenes:ProteinName`) %>%
  tidyr::separate_longer_delim(tosplit, delim = ",") %>%
  tidyr::separate(tosplit, into = c("VF_gene_name", "VF_gene_description"), sep = ": ", fill = "right") %>%
  dplyr::mutate_at(vars(VF_gene_name, VF_gene_description), ~ str_trim(.))

# checking number isolates 
APEC_all_merged_cleaned_long %>%
  select(Sample, VF_gene_name) %>%
  pull(Sample) %>% unique() %>% length() # 796 ok
```

Filter the long table to keep only the virulence genes from Camila
```{r}
in_camila_list <- 
  APEC_all_merged_cleaned_long %>%
  filter(VF_gene_name %in% camila_virulence_genes$Gene_name) %>%
  distinct() 

# nb isolates in Camilas' list

in_camila_list %>%
  select(Sample, VF_gene_name) %>%
  pull(Sample) %>% unique() %>% length() # 783 

dim(in_camila_list)
# isolates where genes were not in Camila's list
APEC_all_merged_cleaned %>%
  anti_join(in_camila_list %>% select(Sample)) %>%
  select(Sample,`Virulence_Finder/VirulenceGenes:ProteinName`) %>%
  write_csv("results/APEC_all_merged_cleaned_genes_not_in_camila_list.csv")
```

> 13 isolates had non of the genes that were in Camila's list  
> 783 isolates had a least one of the gene in Camila's list


```{r}
# control
nb_unique_genes = length(unique(in_camila_list$VF_gene_name))  
nb_unique_genes

length(unique(in_camila_list$Sample))
# 783 --- 
```


Now its in the long format so we can make summary statistics

# Summary statistics for virulence genes

- How many have all genes
```{r}
summary_nb_genes <- 
  in_camila_list %>%
  group_by(Sample) %>%
  #filter(Sample == "2018-01-3168-1-2-1-1") %>%
  #select(Sample, VF_gene_name, VF_gene_description) %>%
  summarise(nb_genes = n()) %>%
  arrange(desc(nb_genes)) 

summary_nb_genes_df <- 
  summary_nb_genes %>%
  ungroup() %>%
  group_by(nb_genes) %>%
  summarise(nb_isolates_with_nb_genes = n()) %>%
  arrange(desc(nb_genes))

# if the results directory does not exist, create it
if (!dir.exists("results")) {
  dir.create("results")
}

summary_nb_genes_df %>% 
  write_csv("results/summary_nb_genes.csv")
```

> 34 samples have all genes in Camila's list
>  the detail of nb of genes in the table results/summary_nb_genes.csv
> 749 isolates do not have all the isolates in Camillas list but have a bit one in the list 
> 13 isolates do not have any genes that were listed in Camila's list

```{r}
# How many isolates have all genes VS other genes
summary_nb_genes_df %>% 
  mutate(group = ifelse(nb_genes == nb_unique_genes, "All genes", "Not all genes")) %>%
  group_by(group) %>%
  summarise(nb_isolates = sum(nb_isolates_with_nb_genes)) %>%
  write_csv("results/summary_nb_genes_groups_totals.csv")
  
```

- long format for the table with genes
```{r}
APEC_all_merged_cleaned_wide <- 
  in_camila_list %>%
  select(-VF_gene_description) %>%
  distinct() %>%
  mutate(present = "Yes") %>%
  pivot_wider(
    names_from = VF_gene_name,
    names_glue = "VF_gene_{VF_gene_name}",
    values_from = present, 
    values_fill = "No"
  ) 

length(unique(APEC_all_merged_cleaned_wide$Sample)) # 783 ok 
glimpse(APEC_all_merged_cleaned_wide)  

write_csv(APEC_all_merged_cleaned_wide,
          "results/APEC_all_merged_cleaned_wide.csv")
```

- creating table with combination of genes 
```{r}
APEC_all_merged_cleaned_combination <- 
  in_camila_list %>%
  arrange(Sample, VF_gene_name) %>%
  group_by(Sample) %>%
  summarise(VF_genes_combination = paste(VF_gene_name, collapse = ","), .groups = "drop")

dim(APEC_all_merged_cleaned_combination) # 783 isolates with combinations of genes

APEC_all_merged_cleaned_combination %>%
  write_csv("results/APEC_all_merged_cleaned_combination_genes_per_isolate.csv")
```

- now we can group the combinations
```{r}
APEC_all_merged_cleaned_combination %>%
  group_by(VF_genes_combination) %>%
  summarise(nb_isolates = n(), .groups = "drop") %>%
  arrange(desc(nb_isolates)) %>%
  write_csv("results/APEC_all_merged_cleaned_combination.csv")
```

- summary STs only (note here all 796 isolates)
```{r}
#dim(APEC_all_merged_cleaned)

APEC_all_merged_cleaned %>%
  rename(ST = `mlst/sequence_type`) %>%
  select(Sample, ST) %>%
  group_by(ST) %>%
  summarise(nb_isolates = n(), .groups = "drop") %>%
  arrange(desc(nb_isolates)) %>%
  write_csv("results/APEC_ST_summary.csv")
```


- check for association with ST for the combinations of genes in Camila's list
```{r}
# The combination per sample  
#dim(APEC_all_merged_cleaned_combination)

# need to be merged to the ST 
APEC_ST_genes_combination <- 
  APEC_all_merged_cleaned_combination %>%
  # left_join to get the ST information
  left_join(APEC_all_merged_cleaned %>% 
              select(Sample, `mlst/sequence_type`),
            by = "Sample") %>%
  rename(ST = `mlst/sequence_type`) %>%
  # contingency table genes combination and ST 
  group_by(VF_genes_combination, ST) %>%
  summarise(nb_isolates = n(), .groups = "drop") %>%
  # dirty way to sort the longest combination first
  mutate(length_combination = str_length(VF_genes_combination)) %>%
  arrange(desc(length_combination))  %>%
  select(-length_combination) 
  
View(APEC_ST_genes_combination)

write_csv(APEC_ST_genes_combination, 
          "results/APEC_ST_genes_combination.csv")  
```


# HERE
Maybe a littple plot of gene combination and ST
- let say we will filter out genes combinations where there are less than 10 isolates

```{r}
graph_data <- 
  APEC_ST_genes_combination %>%
  group_by(VF_genes_combination) %>%
  mutate(sum_gene_combination = sum(nb_isolates)) %>%
  group_by(ST) %>%
  mutate(sum_ST = sum(nb_isolates))
```


```{r}
plot_combination_ST <- 
  graph_data %>%
  filter(sum_gene_combination >= 10) %>%
  ggplot(aes(x = VF_genes_combination, y= nb_isolates, fill = ST)) +
  # rectangle
  geom_col(position = "stack", show.legend = FALSE,  linewidth = 0.2, color = "black") +
  geom_text(
    aes(label = ST), # The label content will be the ST value
    position = position_stack(vjust = 0.5), # Crucial for centering text in stacked bars
    color = "black", # Adjust label color for readability
    size = 1      # Adjust label size as needed
  ) +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  labs(title = "Gene combinations and STs",
       subtitle = "ST are indicated in the bars",
       x = "Gene combinations",
       y = "Number of isolates)",
       color = "ST") +
  theme_bw() 

ggsave("results/plot_combination_ST.png", 
       plot = plot_combination_ST, 
       width = 30, height = 30,
       units = "cm",
       dpi = 300, 
       limitsize = FALSE)
```

Otherway to see: ST and then combination as fill collors:
```{r}
  graph_data %>%
  filter(sum_ST >= 10) %>%
  pull(VF_genes_combination) %>%
  unique() %>% length() # 44 combinations of genes

# so we cant put annotations like that 
plot_ST_combination <- 
  graph_data %>%
  filter(sum_ST >= 10) %>%
  # creating a factor for numbering
  mutate(Factor_combination = factor(VF_genes_combination, 
                                          levels = unique(VF_genes_combination))) %>%
  mutate(Factor_combination = as.character(as.integer(Factor_combination))) %>%
  mutate(VF_genes_combination = 
           paste(Factor_combination, ": ", VF_genes_combination)) %>%
  ggplot(aes(x = ST, y= nb_isolates, fill = VF_genes_combination)) +
  # rectangle
  geom_col(position = "stack", show.legend = T,  linewidth = 0.2, color = "black") +
  geom_text(
    aes(label = Factor_combination), # The label content will be the number of the conbination
    position = position_stack(vjust = 0.5), 
    color = "black", 
    size = 1
  ) +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  labs(title = "ST and Gene combinations",
       subtitle = "Gene combinations identifiers are indicated in the bars",
       x = "ST",
       y = "Number of isolates)",
       color = "Gene combinations") +
  theme_bw() +
  theme(legend.position = "bottom", 
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 6),
    legend.key.size = unit(0.5, "cm")
    ) 


ggsave("results/plot_ST_combination.png", 
       plot = plot_ST_combination, 
       width = 30, height = 30,
       units = "cm",
       dpi = 300, 
       limitsize = FALSE)
```

  
# Other summaries 
This will include the 13 with no results in Camila's list
```{r}
APEC_all_merged_cleaned %>%
  names()
```


```{r}
other_summmaries_df <- 
APEC_all_merged_cleaned %>%
  select(`produsentnr`, `saksnr`)
```

- nb of saksnumber per produsent number
```{r}
antall_sak_per_produsent_df <- 
  other_summmaries_df %>%
  group_by(`produsentnr`) %>%
  # remove the lignes were same sak was registred - allows to count sak
  distinct() %>%
  summarise(antall_sak_per_produsent = n(), .groups = "drop") %>%
  arrange(desc(antall_sak_per_produsent)) 

write_csv(antall_sak_per_produsent_df,  "results/APEC_antall_sak_per_produsent.csv")
```

- nb isolates per farm (produsent number)

```{r}
antall_isolater_per_produsent_df <- 
  other_summmaries_df %>%
  group_by(`produsentnr`) %>%
  # here no filtering -> because the number of rows equals the number of isolates
  summarise(antall_isolater_per_produsent = n(), .groups = "drop") %>%
  arrange(desc(antall_isolater_per_produsent))

antall_isolater_per_produsent_df %>%
  write_csv("results/APEC_antall_isolater_per_produsent.csv")

```


# Summary SOLVEIG (I did not look at those)
 
```{r}
#Solveig trying to summarize MLST and serotype results:
MLST_count <-APEC_all_merged %>%
  select(`mlst/sequence_type`)%>%
  group_by(`mlst/sequence_type`)%>%
  summarise(antall=n())%>%
  arrange(desc(antall))
#29 missing mlst, should try to run these again, perhaps on SAGA?

serotype_count <- APEC_all_merged %>%
  select(`SeroTypeFinder/SeroType`)%>%
  group_by(`SeroTypeFinder/SeroType`)%>%
  summarise(antall=n())%>%
  arrange(desc(antall))

mlst_and_serotype_count <- APEC_all_merged %>%
  select(`mlst/sequence_type`, `SeroTypeFinder/SeroType`)%>%
  group_by(`SeroTypeFinder/SeroType`, `mlst/sequence_type`)%>%
  summarise(antall=n(), .groups = "drop") %>%
  arrange(desc(antall))
```

```{r}
#Solveig check how many isolates per produsentnr
prodnr_count <-APEC_all_merged %>%
  select(`produsentnr`) %>%
  group_by(`produsentnr`) %>%
  summarise(antall=n()) %>%
  arrange(desc(antall))
#several samples from APEC-Seq missing information regarding produsentnr. Try to fix this in an easy way?
#update July 2025: Information regarding produsentnummer available for all isolates.

#Check how many isolates per saksnr
sak_count <- APEC_all_merged %>%
  select(`sak`) %>%
  group_by(`sak`) %>%
  summarise(antall=n()) %>%
  arrange(desc(antall))

#Check sak per produsentnr
sak_prod_count <- APEC_all_merged %>%
  select(`sak`,`produsentnr`) %>%
  group_by(`produsentnr`, `sak`) %>%
  summarise(antall=n(), .groups ="drop") %>%
  arrange(desc(`produsentnr`))
#cannot get this to work properly - would like it to count number of sak per produsentnummer, but it lists produsentnr and sak and how many isolates are from that one sak. How can we check number of sak per produsent?
```

